Type theories as quotient inductive-recursive types

### 2025-02-04 tex/main.tex
Draft the coherence proof in S3.4 and some intro in S4

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself and, at least, partially in the proofs of its properties.
\LT{Strict versus definitional?}
%

@@

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, Agda can be convinced to accept our development, using some experimental flags.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\cite{Cockx2014,Altenkirch2016a}
\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
\subsection{Quotient inductive(-inductive)-recursive types}
\LT{Give an example of QIRT: join list?}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

The laws for substitution composition are stipulated using equality constructors.
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
\end{alignat*}
Substitution composition $\blank;\blank$ acts like a list concatenation, so $\sigma; (\tau, t)$ is intuitively equal to $(\sigma; \tau), [\sigma]\,t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
However, by the type of substitution extension, $[\sigma]\,t$ requires to be of type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead, so we have to transport $[\sigma]\,t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.

@@

To repair the local confluence, we add a simultaneously defined function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
\begin{alignat*}{3}
  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
[ \idS ]_{\Tm}\,t          & \reduce t \\
[ \sigma ; \tau ]_{\Tm}\,t & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
[ \pi_1(\sigma, t) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\,t \\
[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
[ \sigma ]_{\Tm}\,t        & \reduce [ \sigma ]\,t, \quad \text{otherwise}
\end{alignat*}
which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.

@@

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \Tel               & :                   && \Ctx \to \Set                                     \\
  \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
  \Gamma & \dplus \emptytel && \reduce \Gamma \\
  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
Then, we define the multiple lifting $\blank \upuparrows \blank$.
Observe that its type, generalising the type \eqref{eq:type-of-lifting} of a single lifting $\blank\uparrow\blank$, should be
\[
\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
\to \Sub\;(\Gamma \dplus ([\sigma]_{\Tel}\,\Xi))\;(\Delta \dplus \Xi)
\]
asking for telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$, but $[\blank]_{\Tel}\blank$ also needs the multiple lifting for the case $[ \sigma ]_{\Tel}\,(\Xi, A)$ (why?).
Hence, we arrive the conclusion that telescope substitution and the multiple lifting by a telescope need to be defined mutually:
\begin{minipage}{.5\linewidth}
\begin{alignat*}{5}
  & [ \sigma ]_{\Tel}\,\emptytel && \reduce \emptytel \\
  & [ \sigma ]_{\Tel}\,(\Xi, A)  && \reduce [ \sigma ]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\linewidth}
\begin{alignat*}{5}
  & \sigma \upuparrows \emptytel && \reduce \sigma \\
  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi), A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip

To prove the coherence property for the multiple lifting, we would need the properties $(\idS\;\{\Gamma\}) \upuparrows \Xi = \idS\;\{\Gamma \dplus  \Xi\}$ and $(\sigma; \tau) \upuparrows \Xi = \sigma; \tau^+$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.

@@

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & [ \idS   ]_{\Tel}       \,\Xi && \reduce \Xi \\
  & [ \sigma ; \tau ]_{\Tel}\,\Xi && \reduce [ \sigma ]_{\Tel}\,([\tau]_{\Tel}\,\Xi)
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & \idS          \upuparrows \Xi && \reduce \idS \\
  & (\sigma;\tau) \upuparrows \Xi && \reduce (\sigma \upuparrows [ \tau ]_{\Tel} \Xi) ; (\tau \upuparrows \Xi)
\end{alignat*}
\end{minipage}\vskip\belowdisplayskip\noindent
whose confluence property is easy to see.
Now, we proceed with the coherence property for the multiple lifting, telescope substitution, and type substitution altogether.
\begin{proposition}[Coherence]
  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the identities
  \danger
  \begin{align*}
  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
\sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
%    &
%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
  \end{align*}
  hold for any telescope $\Xi$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
\end{proposition}
\begin{proof}
  \LT{
    Consider the following cases 
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  Identities in the firs row are straightforward (not completely trivial, though).
  We shall demonstrate the proof for $[\pi_1(\sigma, t)]_{\Ty}\;A = [\sigma]_{\Ty}\;A$ only.

  By induction on $A$ first then prove other identities by induction on $\Xi$.
  }
\end{proof}

\begin{corollary}[Coherence]
  the identity $[\sigma]_{\Ty} A = [\tau]_{\Ty}A$ holds for every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$.
  \danger
\end{corollary}

@@

\section{The eliminator, the standard model, and the equivalence to the QII definition of type theory}
\LT{%
\begin{enumerate}
  \item Introduce motives and methods for the definition in \Cref{subsec:SC+U+Pi}.
  \item The rule $\mathsf{Elim}_{\Ty}\,([ \sigma ]_{\Ty}\,A) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Ty}(\mathsf{Elim}_{\Ty}\,A)$ is not defined but provable. 
  \item $\mathsf{Elim}_{\Tm}\,([ \sigma ]_{\Tm}\,t) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Tm}(\mathsf{Elim}_{\Tm}\,t)$ is proved.
\end{enumerate}
}

\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}

\section{Formal implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

>>>>

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
\LT{terminology: strict vs.\ definitional vs.\ judgemental?}
%

@@

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\cite{Cockx2014,Altenkirch2016a}
\LT{We use $f\;\vec{x} \reduce \vec{t}$ to indicate a function clause of a definition by overlapping patters.}
\subsection{Quotient inductive(-inductive)-recursive types}
\LT{Give an example of QIRT: join list?}
\LT{How the congruence rule for the identity type can be used in QII(R)T?}
\subsection{Definitions by rewrite rules}
\cite{Cockx2020,Cockx2021}

@@

The laws for substitution composition are stipulated using equality constructors.
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma} && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} && \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] t}}
\end{alignat*}
Substitution composition $\blank;\blank$ acts like a list concatenation, so $\sigma; (\tau, t)$ is intuitively equal to $(\sigma; \tau), [\sigma]\,t$ whereas the term $t$ is of type $\Tm\;\Delta\;([\tau]\;A)$. 
However, by the type of substitution extension, $[\sigma]\,t$ requires to be of type $\Tm\;\Gamma\;([\sigma; \tau]\,A)$ instead, so we have to transport $[\sigma]\,t$ explicitly along $[;]_{\Ty}^{-1}\colon [\sigma]\,[\tau]\;A = [\sigma;\tau]\,A$ to obtain a term of type $\Tm\;\Gamma\;([\sigma;\tau]\;A)$.

@@

To repair the local confluence, we add a simultaneously defined function $[\blank]_{\Tm}\blank$ for term substitution apart from the explicit term substitution $[\blank]\blank$ as follows
\begin{alignat*}{3}
  [\blank]_{\Tm}\blank & : (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\,\Delta\,A \to \Tm\,\Gamma\,([\sigma]_{\Ty}\,A) \\
[ \idS ]_{\Tm}\,t          & \reduce t \\
[ \sigma ; \tau ]_{\Tm}\,t & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
[ \pi_1(\sigma, t) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\,t \\
[ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
[ \sigma ]_{\Tm}\,t        & \reduce [ \sigma ]\,t, \quad \text{otherwise}
\end{alignat*}
which reduces substitutions that occur in \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4} recursively so that the reduction can be propagated from type to term substitution.
The remaining cases reduce to the explicit term substitution $[\sigma]\,t$.
Note that $[\blank]_{\Tm}\blank$ can be defined with the standard semantics for function definition, since there are no overlapping patterns.
\begin{remark}
  The function $[\blank]_{\Tm}\blank$ can be understood as a \emph{smart constructor} for the explicit term substitution $[\blank]\blank$, a well-known idiom in functional programming, since the function $[\blank]_{\Tm}\blank$ (mostly) builds an explicit substitution but computes certain cases away.
\end{remark}

@@

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \Tel               & :                   && \Ctx \to \Set                                     \\
  \emptytel          & : \implicit{\Gamma} && \Tel\;\Gamma \\
  \blank,\blank      & : \implicit{\Gamma} && (\Xi : \Tel\;\Gamma) (A : \Ty\;(\Gamma \dplus \Xi)) \to \Tel\;\Gamma
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  \blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
  \Gamma & \dplus \emptytel && \reduce \Gamma \\
  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
Then, we define the multiple lifting $\blank \upuparrows \blank$.
Observe that its type, generalising the type \eqref{eq:type-of-lifting} of a single lifting $\blank\uparrow\blank$, should be
\[
\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
\to \Sub\;(\Gamma \dplus ([\sigma]_{\Tel}\,\Xi))\;(\Delta \dplus \Xi)
\]
asking for telescope substitution $[\blank]_{\Tel}\blank :  \Sub\;\Gamma\;\Delta \to \Tel\;\Delta \to \Tel\;\Gamma$, but $[\blank]_{\Tel}\blank$ also needs the multiple lifting for the case $[ \sigma ]_{\Tel}\,(\Xi, A)$ (why?).
Hence, we arrive the conclusion that telescope substitution and the multiple lifting by a telescope need to be defined mutually:
\begin{minipage}{.5\linewidth}
\begin{alignat*}{5}
  & [ \sigma ]_{\Tel}\,\emptytel && \reduce \emptytel \\
  & [ \sigma ]_{\Tel}\,(\Xi, A)  && \reduce [ \sigma ]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\linewidth}
\begin{alignat*}{5}
  & \sigma \upuparrows \emptytel && \reduce \sigma \\
  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip

To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows [\tau]_{\Tel} \Xi); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus [\idS]_{\Tel}\,\Xi)\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.

@@

\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & [ \idS   ]_{\Tel}       \,\Xi && \reduce \Xi \\
  & [ \sigma ; \tau ]_{\Tel}\,\Xi && \reduce [ \sigma ]_{\Tel}\,([\tau]_{\Tel}\,\Xi)
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & \idS          \upuparrows \Xi && \reduce \idS \\
  & (\sigma;\tau) \upuparrows \Xi && \reduce (\sigma \upuparrows [ \tau ]_{\Tel} \Xi) ; (\tau \upuparrows \Xi)
\end{alignat*}
\end{minipage}\vskip\belowdisplayskip\noindent
whose confluence property is easy to see.

Now, we proceed with the coherence property for the recursive functions introduced so far.
The first set of equality constructors are easy to prove, since the lifting and substitution operations compute independently for these cases:
\begin{proposition}[Coherence, I]
  \danger
  For the following equality constructors $p : \sigma = \tau$
  \[
    \mathsf{idr} : \sigma;\idS = \sigma, \quad
    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma), 
  \]
  the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
  hold definitionally for any telescope $\Xi$ and any type $A$ and
  \[
    \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}
    \tau   \upuparrows \Xi
  \]
  holds propositionally for any telescope $\Xi$.
\end{proposition}
\begin{proof}
  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds definitionally by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
  Take $\mathsf{idr}$ as an example, we can see 
  \begin{align*}
    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows [\idS]_{\Tel} \Xi) ; (\idS \upuparrows \Xi) \\
                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
                                   & = \sigma \upuparrows \Xi
  \end{align*}
  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.

  For other two equality constructors, the arguments follow similarly.
\end{proof}
The second set of equality constructors are more involved, as we have to prove the coherence for the lifting and substitution operations altogether mutually:
\begin{proposition}[Coherence, II]
  For the following equality constructors $p : \sigma = \tau$
  \begin{alignat*}{5}
    \pi_1\beta    & : \pi_1(\sigma, t)  && = \sigma \qquad\qquad & \mathsf{concat} & : \sigma; (\tau, t) && = (\sigma;\tau), [\sigma]\, t \\
    \emptyctx\eta & : \sigma            && = \emptyctx & \pi\eta & : \sigma && = \pi_1\sigma, \pi_2\sigma
  \end{alignat*}
   the identities
  \danger
  \begin{align*}
  [\sigma]_{\Tel}\,\Xi & = [\tau]_{\Tel}\,\Xi 
                         & [\sigma \upuparrows \Xi ]_{\Ty}\,A & =^{\lam{\gamma}\Ty\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Ty}\,A \\
  \sigma \upuparrows \Xi & =^{\lam{\gamma}\Sub\;(\Gamma\dplus [\gamma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}_{\alert{p}} \tau \upuparrows \Xi
%    &
%  [\sigma \upuparrows \Xi ]_{\Tm}\,t & =^{\lam{\gamma}\Tm\;(\Gamma \dplus [\gamma]_{\Tel}\Xi)\;([\gamma \upuparrows \Xi]_{\Ty} A)}_{\alert{p}} [\tau \upuparrows \Xi ]_{\Tm}\,t
  \end{align*}
  hold for any telescope $\Xi : \Tel\,\Delta$ and type $A : \Ty\;(\Delta\dplus \Xi)$.
\end{proposition}
\begin{proof}
  We demonstrate the proof for $\pi_1\beta: \pi_1(\sigma, t) = t$ and other three cases follow similarly.

  First, we prove the coherence for $[\blank]_{\Ty}\,A$ by induction on $A$, assuming the coherence for the other two identities $[\blank]_{\Tel}\, \Xi$ and $\blank \upuparrows \Xi$, which are then proved by induction on $\Xi$.
  \begin{enumerate}
    \item By induction on $A$, consider the following cases for $[\sigma\upuparrows\Xi]_{\Ty}\,A = [\tau \upuparrows \xi]_{\Ty}\,A$.
      \begin{description}
        \item[$A$ is $\UU$:] By definition
          \[
            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \quad\text{and}\quad [\sigma\upuparrows \Xi]_{\Ty}\,\UU_{\Delta\dplus \Xi} \judgeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}
          \]
          By assumption, $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [ \sigma]_{\Tel}\,\Xi$ and by the congruence $\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi = \Gamma \dplus [ \sigma]_{\Tel}\,\Xi$ holds.
          Then, $\UU_{\Gamma \dplus [\pi_1(\sigma, t)]_{\Tel}\,\Xi} \simeq \UU_{\Gamma \dplus [\sigma]_{\Tel}\,\Xi}$ holds by congruence for $\UU$.
          \LT[noinline]{I think the congruence for $\UU$ is true but not sure how precisely.}

        \item[$A$ is $\El\,u$:]
          Similar to \Cref{prop:coherence-2}, we use \Cref{prop:correctness-3} and the congruence law for $[\blank]\,u$ to derive the desired identity.

        \item[$A$ is $\Pi\,B\,C$:]
          Note that one of the induction hypothesises is
          \[
            [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,C = [\sigma \upuparrows \Xi]_{\Ty}\,C
          \]
          for \emph{arbitrary} $\Xi$, so we can apply the induction hypothesis to $B$ and $C$ below:
          \begin{align*}
            & [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) \\
            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$} \\
            \judgeq{} & \Pi\,([\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,B)\,([(\pi_1(\sigma, t) \upuparrows (\Xi, B)]\,C) && \text{by the definition of $\upuparrows$} \\
            ={}       & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([\sigma \upuparrows (\Xi, B)]\,C) && \text{by the induction hypothesises} \\
            \judgeq{} & \Pi\,([\sigma \upuparrows \Xi]_{\Ty}\,B)\,([(\sigma \upuparrows \Xi) \uparrow B]\,C) && \text{by the definition of $\upuparrows$} \\
            \judgeq{} & [\sigma\upuparrows \Xi]_{\Ty}\,(\Pi\,B\,C) && \text{by the definition of $[\blank]_{\Ty}(\Pi\,B\,C)$.}
          \end{align*}
          Note that 
      \end{description}
    \item By induction on $\Xi$, consider following cases for $[\pi_1(\sigma, t)]_{\Tel}\,\Xi = [\sigma]_{\Tel}\,\Xi$. 
      \begin{description}
        \item[$\Xi$ is the empty telescope $\emptytel$:] By definition, $[\pi_1(\sigma, t)]_{\Tel}\,\emptytel \judgeq \emptytel \judgeq [\sigma]_{\Tel}\,\emptytel$.
        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
          \begin{align*}
            [\pi_1(\sigma, t)]_{\Tel}\,(\Xi, A) & \judgeq [\pi_1(\sigma, t)]_{\Tel}\,\Xi, [\pi_1(\sigma, t) \upuparrows \Xi]_{\Ty}\,A && \text{by definition} \\
                                                & = [\sigma]_{\Tel}\,\Xi, [\sigma \upuparrows \Xi]_{\Ty}\,A \\
                                                & \judgeq [\sigma]_{\Tel}(\Xi, A) && \text{by definition}
          \end{align*}
      \end{description}
    \item By induction on $\Xi$, consider the following cases for $\pi_1(\sigma, t) \upuparrows \Xi = \sigma \upuparrows\,\Xi$.
      \begin{description}
        \item[$\Xi$ is the empty telescope $\emptytel$:] $\pi_1(\sigma, t) \upuparrows \emptytel \judgeq \pi_1(\sigma, t) = \sigma \judgeq \sigma \upuparrows \emptytel$ by definition and the equality constructor.
        \item[$\Xi$ is the telescope extension $\Xi, A$:] 
          \begin{align*}
            \pi_1(\sigma, t) \upuparrows (\Xi, A) & \judgeq (\pi_1(\sigma, t) \upuparrows \Xi) \uparrow A \\
                                                  & =^{} (\pi_1(\sigma, t) \upuparrows \Xi)^{+} \\
                                                  & =^{\lam{\gamma}\Sub\,(\Gamma\dplus[\gamma]_{\Tel}\,\Xi, [\gamma \upuparrows \Xi]_{\Ty}\,A)\,(\Delta \dplus \Xi, A)}_{p} (\sigma\upuparrows \Xi)^{+} \\
                                                  & = (\sigma \upuparrows \Xi) \uparrow A \\
                                                  & \judgeq \sigma \upuparrows (\Xi , A)
          \end{align*}
      \end{description}
  \end{enumerate}

\begin{corollary}[Coherence]
  The identity $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ holds for every equality constructor $p : \sigma = \tau$.
  \danger
\end{corollary}

@@

\section{The eliminator, the standard model, and the equivalence to the QII definition of type theory}
\LT{%
\begin{enumerate}
  \item Introduce motives and methods for the definition in \Cref{subsec:SC+U+Pi}.
  \item The rule $\mathsf{Elim}_{\Ty}\,([ \sigma ]_{\Ty}\,A) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Ty}(\mathsf{Elim}_{\Ty}\,A)$ is not defined but provable. 
  \item $\mathsf{Elim}_{\Tm}\,([ \sigma ]_{\Tm}\,t) = [ \mathsf{Elim}_{\Sub}\,\sigma ]^{A}_{\Tm}(\mathsf{Elim}_{\Tm}\,t)$ is proved.
\end{enumerate}
}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
After giving the QIIR definition of a type theory, we have to define how to \emph{eliminate} types $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ altogether to another types (called \emph{motives}) by functions (called \emph{methods}) subject to equations corresponding to point constructors and equality constructors (\cref{subsec:eliminator}).
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart (\cref{subsec:equivalence}).
